---
title: "Read mitochondrial hit data from SAM files"
author: "Ben Skinner"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, results='hide',message=F, warning=F, echo=F}
library(knitr)
knitr::opts_chunk$set(
	dev = "png",
	dev.args = list(type = "cairo"),
	dpi = 300
)
```

## Load libraries

```{r message=FALSE}
library(dplyr)
library(ggplot2)
# source("https://bioconductor.org/biocLite.R")
# biocLite("GenomicRanges")
# biocLite("ggbio")
library(GenomicRanges)
library(ggbio)
```

## Load the reference genome

```{r}
# Pig genome data
chrnames   = c(1:18, "X", "Y")
chrlengths = c(274330532, 151935994, 132848913, 130910915, 104526007, 170843587, 121844099, 138966237, 139512083, 69359453, 79169978, 61602749, 208334590, 141755446, 140412725, 79944280, 63494081, 55982971, 125939595, 43547828)
names(chrlengths) = chrnames
ref.genome = GRanges(chrnames, IRanges(1, chrlengths))
```

```{r}
breeds = c("Duroc", "Berkshire", "Landrace")
```


## Find shared insertions in breed data

In order to combine hits within a general region, the read coordinates are extended by 16kb in either direction.

The goal here is to combine all sequence intervals across all animals into a single list of unique regions of the genome in which MT pairs were found. The number of animals with a hit in each of these regions is then counted. Only regions occurring in at least 60% of the animals is considered a possible 'real' hit.

```{r}
findSharedInsertions = function(breed){
  
  cat("Running on", breed,"\n")
  
  ## Load data files and reference genome
  dir = paste0("Y:/bms41/Pigs/NuMTs/filtered_reads/",breed,"/")
  files = list.files(dir, pattern=".*.filt.MT.sam", all.files=F, full.names=T)
  
  ## Read the files to sequence objects
  readSamFile = function(file){
    
    
    data = tryCatch({
        read.csv(file, skip=1, header=F, stringsAsFactors = F, sep="\t")
      }, error=function(cond){
        message(paste("Cannot read ", file))
        message("Original error message:")
        message(cond)
        return(NA)
      })
    locs = data %>% select("Flag"=V2, V3, V4, V7, V8, V10) %>% filter(V3=="MT"| V7=="MT")
    
    r1 = locs %>% 
      filter(V3=="MT") %>% 
      select(Flag, "MT_pos" = V4,"Chr"=V7,"Chr_pos"=V8, "Seq"=V10  ) %>%
      mutate("ChrStrand" = ifelse(Flag&32, "-", "+"))  # flag 32 for next read is reverse complemented
    
    
    
    r2 = locs %>% filter(V3!="MT") %>% select(Flag, "MT_pos" = V8,"Chr"=V3,"Chr_pos"=V4, "Seq"=V10  ) %>%
      mutate("ChrStrand" = ifelse(Flag&16, "-", "+")) # flag 16 for first read is reverse complemented
    
    mge = rbind(r1, r2) %>% mutate(TLen = nchar(Seq)) %>% select(-Seq)
    

    

    
    min.reads = 10
    all.reads = with(mge, GRanges(Chr, IRanges(Chr_pos, Chr_pos+TLen)))
    seqlevels(all.reads) = chrnames
    seqlengths(all.reads) = chrlengths
    
    # Combine to overlapping regions
    seqs = reduce(all.reads)
    # Find the number of reads in each unique range within the sample
    cat("Testing", length(seqs),"unique ranges; ")
    chr.reads = split(all.reads, seqnames(all.reads))

    findReadsWithCoverage = function(i){
      combined.range = seqs[i]
      chr = unique(seqnames(combined.range))
      hits = sum(countOverlaps(chr.reads[[chr]], combined.range))
      return(hits>min.reads)
    }
    
    valid.reads = lapply(1:length(seqs), findReadsWithCoverage)
    
    
    final.reads = seqs[valid.reads==TRUE]
    
    plotGr = function(seqname){
      g = ggbio()+geom_rect(final.reads[seqname], main=paste(file, seqname))
      ggsave(file=paste0(file, "_", seqname, ".png"))
    }
    lapply(unique(seqnames(final.reads)), plotGr)
    
    cat("Found", length(final.reads),"ranges with at least",min.reads,"reads\n")
    seqlevels(final.reads) = chrnames
    seqlengths(final.reads) = chrlengths
    final.reads
  }
  
  seqs = lapply(files, readSamFile)
  
  seqs = na.omit(seqs)
  
  
  # Calculate overlaps
  uniqueGranges = function(list){
    result = list[[1]]
    
    for(ranges in list){
      
      result = c(result, ranges)
    }
    result = reduce(result)
    cat("Found", length(result),"unique ranges\n")
    result
  }
  
  unique.ranges = uniqueGranges(seqs)
  
  presentInAtLeast = function(list, unique.ranges, percent){
    
    min.overlaps = (length(list) * percent)/100
    cat("Looking for at least", min.overlaps, "overlaps for each range\n")
    
    countOverlappingGRanges = function(test.range){countOverlaps(unique.ranges, test.range ) }
    
    # Find the overlapping ranges
    overlaps = lapply(list, countOverlappingGRanges)
    
    processList = function(list){
      count = rep(0, length(list[[1]]))
      
      for(l in list){
        count = mapply(sum, count, l)
      }
      count
    }
    
    # count the overlapping ranges
    totals = processList(overlaps)
    
    # filter by count
    filterCounts = function(count){ count>min.overlaps }
    
    keeps = filterCounts(totals)
    
    # Create a result variable by subsetting the first range in the list
    result = unique.ranges[1]
    
    for(i in 1:length(unique.ranges)){
      if(keeps[[i]]){
        result = c(result, unique.ranges[i])
      }
    }
    # First row was a placeholder, remove unless in the keep list
    if(!keeps[[1]]){ result = result[2:length(result)]  }
    cat("Found", length(result),"ranges in at least", percent,"% of the samples\n")
    result
  }
  
  shared.ranges = presentInAtLeast(seqs, unique.ranges, 60)
  shared.ranges
}

inserts = lapply(breeds, findSharedInsertions)

```

## Plot

Plot each of the breed insert data

```{r warning=FALSE}
g = ggbio() + circle(ref.genome, geom = "ideo", fill = "gray70")
for(s in inserts){
  g = g + circle(s, geom = "rect", color = "steelblue", size=1)
}
g = g +
  # circle(shared.ranges, geom = "rect", color = "red", size=2) +
  circle(ref.genome, geom = "scale", size = 2)+
  circle(ref.genome, geom = "text", aes(label = seqnames), vjust = 0, size = 3)
g

```

## Data

Visualise the aggregate hits

```{r results='asis'}

for(s in inserts){
  df = data.frame(chr=seqnames(s),
                  start=start(s)-1L,
                  end=end(s))
  
  print(kable(df, caption = "MT hit table"))
}

```


## Plot individual reads per animal


```{r fig.height=3, fig.show='asis'}

plotAnimalInsertions = function(breed){
  
  cat("Running on", breed,"\n")
  
  ## Load data files and reference genome
  dir = paste0("Y:/bms41/Pigs/NuMTs/filtered_reads/",breed,"/")
  files = list.files(dir, pattern=".*.filt.MT.sam", all.files=F, full.names=T)
  
  ## Read the files to sequence objects
  readSamFile = function(file){
    
    cat("Reading", file,"\n")
    data = read.csv(file, skip=1, header=F, stringsAsFactors = F, sep="\t")
    locs = data %>% select(V3, V4, V7, V8, V10) %>% filter(V3=="MT"| V7=="MT")
    
    r1 = locs %>% filter(V3=="MT") %>% select("MT_pos" = V4,"Chr"=V7,"Chr_pos"=V8, "Seq"=V10  )
    r2 = locs %>% filter(V3!="MT") %>% select("MT_pos" = V8,"Chr"=V3,"Chr_pos"=V4, "Seq"=V10  )
    
    mge = rbind(r1, r2) %>% mutate(TLen = nchar(Seq)) %>% select(-Seq)
    
    reads = with( mge, GRanges(Chr, IRanges( Chr_pos, Chr_pos+TLen)) )
    seqlevels(reads) = chrnames
    seqlengths(reads) = chrlengths
    grl = split(reads, seqnames(reads), drop=T) #split by chr
    
    plotGr = function(seqname){
      ggbio()+geom_rect(grl[seqname], main=paste(file, seqname))
    }
    plots = lapply(unique(seqnames(grl)), plotGr)
    
    for(p in plots){
      print(p)
    }
  }
  
  # lapply(files, readSamFile)
  
    
}

plotAnimalInsertions(breeds)

```
